### 2025-01-31 products/manuscript/Manuscript.qmd
commit to the changes in all those files
especially code and qmd

---
title: "Example Manuscript Template for a Data Analysis Project"
author: "Doreen Kibuule Kalembe"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project). Adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

**Authors**

* First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678) 
* Second Author$^{3, *}$
* Third Autor$^{3, \dagger}$
* Last Author$^{1, 2, \land}$

**Author affiliations**

1. College of Public Health, University of Georgia, Athens, GA, USA.
2. Another Affiliation, Athens, GA, USA.
3. Yet another Affiliation.

$*$ These authors contributed equally to this work.

$\land$ Corresponding author: some@email.com

$\dagger$ Disclaimer: The opinions expressed in this article are the
author's own and don't reflect their employer.

{{< pagebreak >}}

# Summary/Abstract
_Write a summary of your project._
MADA PROJECT PART 1:

TITLE: Sex-based analysis/Differences in TB Immune Activation.

This project is aimed at looking at the differences in the Tuberculosis immune activation between males and females.( Sex-based analysis/Differences in TB Immune Activation
Research Question)

we shall use data from a tuberculosis public repository.
•	Data source: Tidyverse 
https://dataverse.unc.edu/dataset.xhtml?persistentId=doi:10.15139/S3/AYOFEU
with a Sample size of 60 individuals. The data includes 60 different individual that accepted to take part in the study.

The key variables that we shall look at are;

Key variables: 
- CD4 Immune activation count
-	Disease severity indicators(Extent of lung disease andNumber of lungs involved)
-Fat in kgs
- Sex/Gender
- BMI
- Age of the participant 
```{r}
#load packages
library("readxl")
library("readr")
library("dplyr")
library("ggplot2")
library("tidyverse")
library("here")

# follow through with the code folder to see the #variables in the data and how i selected the #variables make this data set. 
#location of the data
data_location <- here::here("data","processed-data","Tb_immune_data.rds")

#loading the data
data<- read_rds(data_location)

#summary of the data
summary(data)
```

Proposed Analysis Method

•	Compare immune activation levels between males and females

•	Correlate immune markers with disease severity measures/Correlation between activation levels and disease severity

•	Use linear regression to analyze relationships between variables

•	Control for potential confounding factors

These are some of our Expected Outcomes.

•	Check and Identify any sex-based differences in immune response.

•	Quantify the relationship between immune activation and disease severity.

•	Propose any insights for sex-specific treatment approaches

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2025-01-31 code/analysis-code/statistical-analysis.R

@@ -1,56 +1,34 @@
-###############################
-# analysis script
-#
-#this script loads the processed, cleaned data, does a simple analysis
-#and saves the results to the results folder
+#load packages
+library("readxl")
+library("readr")
+library("dplyr")
+library("ggplot2")
+library("tidyverse")
+library("here")
+# load the data sets
+Immunology_lab_results <- read_excel("data/raw-data/Immunology lab results.xlsx")
+View(Immunology_lab_results)
+GENDER_DEUTERIUM_DILUTION_DATA <- read_excel("data/raw-data/GENDER DEUTERIUM DILUTION DATA.xlsx")
+View(GENDER_DEUTERIUM_DILUTION_DATA)
 
-#load needed packages. make sure they are installed.
-library(ggplot2) #for plotting
-library(broom) #for cleaning up output from lm()
-library(here) #for data loading/saving
+#select variables in both the data sets and later merge them together.
 
-#path to data
-#note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","processeddata.rds")
+gender_data<-GENDER_DEUTERIUM_DILUTION_DATA%>%
+  select(`Participant ID`, Sex, `Participant age`,BMI, Dose,`Fat in kg`)
+gender_data<-gender_data%>%
+  rename("ID"= `Participant ID`)
 
-#load data. 
-mydata <- readRDS(data_location)
+#select variables you will need in the next data set
+immune_data<-Immunology_lab_results%>%
+  select(ID,`CD4+`, `CD4 Immune activation count`)
 
+#we can now merge the two data sets int one.
+Tb_immune_data<-full_join(gender_data, immune_data,by = "ID")
 
-######################################
-#Data fitting/statistical analysis
-######################################
+#let me first save this data set. i will save this data in the processed data
+# folder since this is the data that i have cleaned and plan on using through
+# out the annalysis.
 
-############################
-#### First model fit
-# fit linear model using height as outcome, weight as predictor
+save_data_location <- here::here("data","processed-data","Tb_immune_data.rds")
+saveRDS(Tb_immune_data, file = save_data_location)
 
-lmfit1 <- lm(Height ~ Weight, mydata)  
-
-# place results from fit into a data frame with the tidy function
-lmtable1 <- broom::tidy(lmfit1)
-
-#look at fit results
-print(lmtable1)
-
-# save fit results table  
[…]

### 2025-02-21 products/manuscript/Manuscript.qmd
commit all changes to the files

@@
 _Provide enough background on your topic that others can understand the why and how of your analysis_ 
+Tuberculosis (TB) remains a significant global public health challenge.
+According to the World Health Organization (WHO), reports show that  10 million new cases and 1.4 million deaths in 2023 [@Santos2007] were as a result of Tuberculosis.
+The spread of M.
+tuberculosis, as the pathogen of TB, has long been hypothesized to occur more often in the household than in the community[@Martinez2017].
+While most of the M.Tuberculosis  transmission in the community is attributed to the unknown contact networks that occurs between individuals, this accounts upto 90% of the transmission.
+The recent outbreak of Tuberculosis in the united states clearly explains the contact network phenomena.
+Overall, TB accounts for one quarter of the annual deaths that happen worldwide.
+An estimated 10.0 million people fell ill with TB in 2018.
+Geographically, Southeast Asia ranked highest (44%), followed by Africa (24%) and the Western Pacific (18%), with lesser percentages in the Eastern Mediterranean (8%), the Americas (3%), and Europe (3%).
+Among women of reproductive years, TB is the largest cause of death [@Dye1998].
+Global Strategies made by WHO, aim at ending the TB epidemic, with targets to reduce the number of deaths caused by TB  by 95% and to decrease new cases by 90% from 2015 to 2035[@Murgia2024].
+
+Mycobacterium tuberculosis infection is acquired mostly through sharing air space with an individual who has active tuberculosis and inhaling droplet aerosols produced by that person.
+Since this bacterium is transmitted via the respiratory route to susceptible individuals in the community, the household of an infectious person is an environment for intense transmission of M.
+tuberculosis where most new cases of infection and diseases might be detected and treated.
+According some reports,they suggest that there are differences in the  responses to tuberculosis between men and women.
+These reports highlight greater barrier to early detection and treatment of tuberculosis to be greater for women than men.
+
+The World Health Organisations, among others, attributed to the differences in the tuberculosis trends between men and women to access to care and treatment, pregnancy effects and gender rules that diminish the social capital of women hence lowering incidence among them[@CCWhalen2023].
+Not forgetting that biological mechanisms might influence these differences,  previous studies show that men have a higher incidence and prevalence of disease and this might lead to bad treatment outcomes as compared to women.
+
+Since the balance of fat and lean tissue differ in men and women especially at the time of presentation, the objective of this study will be to find out whether the more immune activation there is, the more infected / sicker is the person.
 
 ## Description of data and data source
+The data we used for this project was got from the TRAC dataverse and is fully accessible to everyone.
+This is considered a safe place where the the Tuberculosis data from the TRAc project is kept and can be used by anyone who wants contribute to TB research.
+More details about the data can be found [here]!(https://dataverse.unc.edu/dataset.xhtml?persistentId=doi:10.15139/S3/AYOFEU).
+The data is publicaly available and can be accessed by anyone.
 
@@
 _State the research questions you plan to answer with this analysis._
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
+Since the balance of fat and lean tissue differ in men and women especially at the time of tuberculosis presentation, the objective of this study will be to find out whether there are differences in the Tuberculosis immune activation between males and females.
+We shall hypothesize on the more immune activation there is, the more infected / sicker the person.
 
@@
 _Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
+Study Design
+This study utilized a cohort design analyzing data from 2021.
+We included all patients that were 15 years and older.
+The candidate must have had TB as their first episode, not on any hormonal therapy and does not take any hormonal birth control.
+The participant was not supposed to have diabetes melitus, cancer or HIV.Similarly,the exclusion criteria applied.
+All participants that were below the age of 15, reported TB as their second episode, took any form of hormanl therapy or biirth control , had diabetes melitus, cancer or HIV were excluded from the study.
+The variables of interest were extracted from the TRAC dataset  and exported to an MS Excel sheet.
+The outcome variable was CD4+ immune activatin counts.
+The primary independent variable was sex (cis-gender) and this was categorized as male and female.
+The covariates were age at diagnosis,BMI, Fat in Kg, lean body mass in kg(LBM in kg), CD4+ and CD4 immune activation counts.
 
 ## Schematic of workflow
-
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
+ Chromosomal sex and hormones play a role in the modulation of immune responses to Mtb and thereby contribute to the general outcome of disease.
+The first cells that encounter Mtb are known as Macrophages.
+These initiate a local inflammatory reaction which may differ between the sexes[@Hertz2018].
+This causes the long term containment of TB infection to differ between men and women since their adaptive immune responses are different.
+@tbfig-1
 
 ```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
+#| label: fig-1
+#| fig-cap: "Impact of sex differences on the immune response to Mtb and their influence disease outcome."
 #| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
+knitr::include_graphics(here("assets","tbfig-1.png"))
 ```
-
-## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
 
@@
 All materials, including files that contain code, should be commented well so everyone can follow along._
+
+We imported the data from MS Excel to R.
+We explored the data and found some missing values for some variables.
+We decided to drop one observation since the person lacked the primary exposure and did not have most of the details we needed for our final analysis.
+We categorized the participants age into four categories.
+That is 17-<28, 28-<40, 40-<60 and >=60.
 
@@
 Even more can be in the R and Quarto files that are part of your project._
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
+@tbl-TABLE1 shows a summary of the data
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
+#| label: tbl-TABLE1
+#| tbl-cap: "Data descriptive summary table."
 #| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
+resulttable=readRDS("../../results/tables/TABLE1.rds")
 knitr::kable(resulttable)
@@
 
-@fig-result shows a scatterplot figure produced by one of the R scripts.
+@fig1-result shows a scatter plot of lean body mass in kg and CD4
 
 ```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
+#| label: fig1-result
+#| fig-cap: "Scatter plot of CD4+ cells versus LBM in kg."
 #| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
+knitr::include_graphics(here("results","figures","scatterplot1.png"))
+```
+
+@fig2-result shows a scatter plot of CD4 vs Fat in kg.
+```{r}
+#| label: fig2-result
+#| fig-cap: "Scatter plot of CD4+ cells versus Fat in kg."
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","scatterplot2.png"))
+```
+
+@plot1-result a boxplot of the cd4 immune activation count by sex.
+```{r}
+#| label: plot1-result
+#| fig-cap: "Box plot to show the distribution of cd4 count immune activation counts among males and females."
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","boxplot1.png"))
+```
+
+@plot2-result a boxplot to show the distributions of fat in kilograms among males females.
+```{r}
+#| label: plot2-result
+#| fig-cap: "Box plot showing Fat in kg among males and females."
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","boxplot2.png"))
 ```

### 2025-02-21 code/analysis-code/statistical-analysis.R

@@ -1,4 +1,5 @@
 #load packages
+install.packages("table1")
 library("readxl")
 library("readr")
 library("dplyr")
@@ -11,10 +12,15 @@ View(Immunology_lab_results)
 GENDER_DEUTERIUM_DILUTION_DATA <- read_excel("data/raw-data/GENDER DEUTERIUM DILUTION DATA.xlsx")
 View(GENDER_DEUTERIUM_DILUTION_DATA)
 
+##data cleaning.
+
+
 #select variables in both the data sets and later merge them together.
 
 gender_data<-GENDER_DEUTERIUM_DILUTION_DATA%>%
-  select(`Participant ID`, Sex, `Participant age`,BMI, Dose,`Fat in kg`)
+  select(`Participant ID`, Sex, `Participant age`,BMI, Dose,`Fat in kg`,`LBM in kg`,
+          `Fat %`, `LBM in %`)
+         
 gender_data<-gender_data%>%
   rename("ID"= `Participant ID`)
 
@@ -32,3 +38,175 @@ Tb_immune_data<-full_join(gender_data, immune_data,by = "ID")
 save_data_location <- here::here("data","processed-data","Tb_immune_data.rds")
 saveRDS(Tb_immune_data, file = save_data_location)
 
+#dropping the last observation from the dataset since the last observation
+#doesnt make sense to our annalysis 
+#loading the data
+data<- read_rds(data_location)
+
+# dropping the last observation in the dataset since it makes no sense
+# to our annalysis and also dropping the variable dose since we dont really need it.
+data<- data[-61,-5] #dropped already. 
+
+#Lets categorise age so we can have a clear picture of what is happening.
+#first lets look at its 
+summary(data$`Participant age`)
+Final_data <- data %>%
+  mutate(Age_cat = case_when(
+    `Participant age` >= 17 & `Participant age` < 28 ~ 0,
+    `Participant age` >= 28 & `Participant age` < 40 ~ 1,
+    `Participant age` >= 40 & `Participant age` < 60 ~ 2,
+    `Participant age` >= 60  ~ 3,  # for all graeter than 60
+    TRUE ~ NA_real_  # Handles missing or unexpected values
+  ))
+
+
+
+##eXPLORATION OF OUR DATA.
+#summary of the data
+summary(Final_data)
+#mean age is of the participants is 30, minimum age is 17 while the maax is 70
+
[…]

### 2025-02-21 code/analysis-code/statistical-analysis.R
commit to changes in the file

@@ -108,7 +108,7 @@ p1<-ggplot(Final_data, aes(x=`LBM in kg`
 plot(p1)
 
 figure_file = here("results","figures","Scatterplot1.png")
-ggsave(filename = figure_file, plot=p1, width = 7, height = 5, dpi = 300)
+ggsave(filename = figure_file, plot=p1, width = 5, height = 3.7, dpi = 300)
 
 
 ##considering immune counts
@@ -128,7 +128,7 @@ p2<-ggplot(Final_data, aes(x= `Fat in kg`
   theme_minimal()
 plot(p2)
 figure_file = here("results","figures","Scatterplot2.png")
-ggsave(filename = figure_file, plot=p2, width = 7, height = 5, dpi = 300)
+ggsave(filename = figure_file, plot=p2, width = 5, height = 3.7, dpi = 300)
 ##immune activation counts
 ggplot(Final_data, aes(x= `Fat in kg`
                  , y=`CD4 Immune activation count`)) + 
@@ -145,7 +145,7 @@ p3<-ggplot(Final_data, aes(x = factor(Sex), y = `CD4 Immune activation count`,
   theme_minimal()
 plot(p3)
 figure_file = here("results","figures","boxplot1.png")
-ggsave(filename = figure_file, plot=p3, width = 7, height = 5, dpi = 300)
+ggsave(filename = figure_file, plot=p3, width = 5, height = 4, dpi = 300)
 
 
 
@@ -155,7 +155,7 @@ p4<-ggplot(Final_data, aes(x = Sex, y = `Fat in kg`, fill = Sex)) +
   ggtitle("Fat in kg by Sex")
 plot(p4)
 figure_file = here("results","figures","boxplot2.png")
-ggsave(filename = figure_file, plot=p4, width = 7, height = 5, dpi = 300)
+ggsave(filename = figure_file, plot=p4, width = 5, height = 3.8, dpi = 300)
 
 ##Mean cd4 count by sex
 ggplot(Final_data, aes(x = factor(Sex), y = `CD4 Immune activation count`, fill = factor(Sex))) +

### 2025-02-22 products/manuscript/Manuscript.qmd
commit changes to the files
commit to the changes

@@
 # Summary/Abstract
-_Write a summary of your project._
+
 MADA PROJECT PART 1:
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_ 
+
 Tuberculosis (TB) remains a significant global public health challenge.
@@
 ## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
+
 Since the balance of fat and lean tissue differ in men and women especially at the time of tuberculosis presentation, the objective of this study will be to find out whether there are differences in the Tuberculosis immune activation between males and females.
@@
 
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
-
 {{< pagebreak >}}

### 2025-03-14 products/manuscript/Manuscript.qmd
commit changes  to all files.

@@
 # Methods 
+Study Design
 
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
-Study Design
 This study utilized a cohort design analyzing data from 2021.
@@
 All participants that were below the age of 15, reported TB as their second episode, took any form of hormanl therapy or biirth control , had diabetes melitus, cancer or HIV were excluded from the study.
+
 The variables of interest were extracted from the TRAC dataset  and exported to an MS Excel sheet.
@@
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
 
@@
 ## Statistical analysis
-_Explain anything related to your statistical analyses._
+
+The final cleaned data set contains some negative fat values.
+Negative values occur when the deuterium dose has not had sufficient time to fully equilibrate with body water, or the dose was not completely consumed.
+This is seen as low deuterium enrichment, resulting in an overestimation of the size of the body water pool and hence high FFM (FFMI = LBM / height2)and low % body fat.
 
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
+We explored the data and using figures and summary tables.
+We used box plots, histograms and and correlation plots to see whether we can find something interesting in the data.
+Scatter plots were used to see the trends in the data and also see whether there are some interesting relationships between the continuous variables in the data.
+
 @tbl-TABLE1 shows a summary of the data
@@
 ## Basic statistical analysis
-
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
+We used simple linear model to inspect whether there was an association between our main outcome CD4 immune activation count and the the primary predictor sex.
+To improve the model, we added in other predictors that could influence the immune activation count by performing a multi-variant linear regression anlaysis.
+This was done make sure we minimize confounding.
 
@@
 ## Full analysis
+In order to take care of confounding, we performed a multi-regression analysis and the results of the analysis are summarized in the table below.
 
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
+This @tbl-model_table2 shows a summary of a  multi-variant linear regression model fit.
+```{r}
+#| label: tbl-model_table2
+#| tbl-cap: "Linear model summary table 2."
+#| echo: FALSE
+model_table2= readRDS(here("results","tables","model_table2.rds"))
+knitr::kable(model_table2)
+```
 
-Example @tbl-resulttable2 shows a summary of a linear model fit.
+We went throught the steps of model building.
+First, we tested the null model which comprised of our main outcome and sex as the predictor.
+We did a stepwise selection using a foward direction approach with a our full model with all the interaction effects.
+We found that adding the interaction terms did not improve the model performance.
+We then considered the null model in comparison with the full model but this time round without interaction effects but all the predictors.
+Using the same procedure of stepwise selection using the foward direction chose the a simple linear model with just cd4+ as the only additional predictor.
+We performed a simple linear regression on the data between the CD4 immune activation count as the main outcome and sex plus cd4+ as predictors. We considered females  to be our reference level.
 
+This @tbl-model_table3 shows a summary of a  final linear model fit.
 ```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
+#| label: tbl-model_table3
+#| tbl-cap: "Linear model fit table 3."
 #| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
+model_table3 = readRDS(here("results","tables","model_table3.rds"))
+knitr::kable(model_table3)
 ```
@@
 # Discussion
+
+Our findings highlight a notable difference in CD4 immune activation count rates between males and females, with males exhibiting higher risk of TB infection and disease than females.
+The results show that men have a 735.34 increased immune activation count as compared to women.
+This result aligns with previous studies that have reported sex-based disparities in health outcomes and mortality risks as a result of TB[@Hertz2018],[@Mara2018].
+Some studies still attribute the increased risk among males to be as a result of biological differences or differences in the accessibility of resources, especially healthcare.
+This is generally because women have better prognosis than men and they have better health seeking behavior as compared to their male counterparts.
+
+Research shows that most women have higher fat repositories than men[@Karastergiou2012] and some studies show that the higher the fat percentage an individual has, the lower the risk of progression of the tuberculosis infection to disease hence lowering mortality rates among females compared to men.
+In addition, women tend to have more lean body mass compared to men hence having a lower chance of suffering from adverse effects of TB compared to men.
 

### 2025-03-14 code/analysis-code/statistical-analysis.R

@@ -1,5 +1,5 @@
 #load packages
-install.packages("table1")
+#install.packages("table1")
 library("readxl")
 library("readr")
 library("dplyr")
@@ -208,5 +208,88 @@ t.test(`Fat in kg`~ Sex, data = Final_data)
   #modify_spanning_header(c("stat_1", "stat_2") ~ "**Sex Groups**") 
 #bold_labels()
 
+#simple models with one predictor.
+#since sex is our major exposure of interst, we shall consider it first and then 
+#see how our outcome performs with it. we shall first see how this goes.
+model1 <- lm(`CD4+` ~ relevel(factor(Sex), ref = "F"), data = Final_data)
+summary(model1)
+#lets try it with the males as our reference group
+model2 <- lm(`CD4+` ~ relevel(factor(Sex), ref = "M"), data = Final_data)
+summary(model2)
 
-  
+#Using cd4 immune activation count .
+model3 <- lm(`CD4 Immune activation count` ~ relevel(factor(Sex), ref = "F"), data = Final_data)
+
+model_summary <- summary(model3)
+model_summary
+model_table1 <- as.data.frame(model_summary$coefficients)
+knitr::kable(model_table1)
+
+
+save_table2<-here::here("results","tables", "model_table1.rds")
+saveRDS(model_table1,file =save_table2 )
+
+#lets try to improve the model and also control for confounders. 
+
+
+model4 <- lm(`CD4 Immune activation count`~ relevel(factor(Sex), ref = "F") +
+               `Participant age` + BMI + `LBM in kg` + `Fat in kg` + `CD4+`,
+            data = Final_data)
+ summary(model4)
+model_summary2 <- summary(model4)
+model_table2 <- as.data.frame(model_summary2$coefficients)
+knitr::kable(model_table2)
+
+
+save_table3<-here::here("results","tables", "model_table2.rds")
+saveRDS(model_table2,file =save_table3 )
+#since we have our full model, lets try to perform a step wise selection procedure
+#to see which model will perform better.
+
+# First, fit your full model with all predictors so that we can see whether it is 
+#relevant for us to have all these predictors or not.
+model4 <- lm(`CD4 Immune activation count` ~ relevel(factor(Sex), ref = "F") + 
+                   `Participant age` + BMI + `LBM in kg` + `Fat in kg` + `CD4+`,
+                 data = Final_data)
+#then lets have a null model
+# Create the null model (intercept only)
[…]

### 2025-03-14 products/manuscript/Manuscript.qmd
commit to all changes to the manuscript

@@
 ---
-
-The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project). Adjust as needed. 
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
 
@@
 1. College of Public Health, University of Georgia, Athens, GA, USA.
-2. Another Affiliation, Athens, GA, USA.
-3. Yet another Affiliation.
 
@@
 
-$\land$ Corresponding author: some@email.com
+$\land$ Corresponding author: kelvinbwambale8@gmail.com
 
@@
 # Summary/Abstract
-
-MADA PROJECT PART 1:
 

### 2025-03-24 products/manuscript/Manuscript.qmd
commit all changes to these files
commit to the changes made on the manuscript and the
code

@@
 * Second Author$^{3, *}$
-* Third Autor$^{3, \dagger}$
-* Last Author$^{1, 2, \land}$
 
@@
 
-Proposed Analysis Method
+Proposed Analysis Method.
 
@@
 
-Mycobacterium tuberculosis infection is acquired mostly through sharing air space with an individual who has active tuberculosis and inhaling droplet aerosols produced by that person.
-Since this bacterium is transmitted via the respiratory route to susceptible individuals in the community, the household of an infectious person is an environment for intense transmission of M.
-tuberculosis where most new cases of infection and diseases might be detected and treated.
-According some reports,they suggest that there are differences in the  responses to tuberculosis between men and women.
-These reports highlight greater barrier to early detection and treatment of tuberculosis to be greater for women than men.
+Mycobacterium tuberculosis infection primarily spreads when a person shares airspace with someone who has active tuberculosis and inhales infectious droplet aerosols.
+Since the bacterium transmits through respiratory routes to susceptible individuals, the household of an infected person becomes a critical environment for intense M.
+tuberculosis transmission, where most new infections and cases can be detected and treated.
+Research reports indicate significant differences in tuberculosis responses between men and women.
+Notably, these studies suggest that women face greater barriers to early detection and treatment of tuberculosis compared to men, potentially impacting disease outcomes and progression.
 
-The World Health Organisations, among others, attributed to the differences in the tuberculosis trends between men and women to access to care and treatment, pregnancy effects and gender rules that diminish the social capital of women hence lowering incidence among them[@CCWhalen2023].
+The World Health Organisations, among others, attributed to the differences in the tuberculosis trends between men and women to access to care and treatment, pregnancy effects,body fat mass and gender rules that diminish the social capital of women hence lowering incidence among them[@CCWhalen2023].
 Not forgetting that biological mechanisms might influence these differences,  previous studies show that men have a higher incidence and prevalence of disease and this might lead to bad treatment outcomes as compared to women.
 
-Since the balance of fat and lean tissue differ in men and women especially at the time of presentation, the objective of this study will be to find out whether the more immune activation there is, the more infected / sicker is the person.
+Body fat plays an important role in supporting immune function, and women typically maintain higher fat storage than men.
+This increased fat storage may contribute to women's stronger immune responses.
+Since the balance of fat and lean tissue differs significantly between men and women, particularly at the time of disease presentation, the objective of this study is to determine whether higher immune activation correlates with greater infection severity or illness.
 
@@
 
-This study utilized a cohort design analyzing data from 2021.
-We included all patients that were 15 years and older.
-The candidate must have had TB as their first episode, not on any hormonal therapy and does not take any hormonal birth control.
-The participant was not supposed to have diabetes melitus, cancer or HIV.Similarly,the exclusion criteria applied.
-All participants that were below the age of 15, reported TB as their second episode, took any form of hormanl therapy or biirth control , had diabetes melitus, cancer or HIV were excluded from the study.
+This study utilized a cross sectional study design analyzing data from 2017.Sixty TB-confirmed participants (30 males, 30 females, aged 15+) were recruited from Mulago Hospital.
+
+\textbf{inclusion and exclusion critera}
+
+Our study included patients aged 15 years and older who were experiencing their first episode of tuberculosis.
+To maintain data integrity, we excluded individuals who were receiving hormonal therapy or using hormonal contraceptives.
+Additionally, patients with comorbidities such as diabetes mellitus, cancer, or HIV were excluded from the study.
+Conversely, we applied strict exclusion criteria to ensure a focused research population: all individuals under 15 years of age, those with recurrent TB episodes, patients taking any form of hormonal therapy or contraception, and those diagnosed with diabetes mellitus, cancer, or HIV were not enrolled in the study.
 
@@
 The covariates were age at diagnosis,BMI, Fat in Kg, lean body mass in kg(LBM in kg), CD4+ and CD4 immune activation counts.
+other variables  like fat mass index (FMI) and fat-free mass index (FFMI) were also obtained.
 

### 2025-03-24 code/analysis-code/statistical-analysis.R

@@ -1,11 +1,15 @@
 #load packages
 #install.packages("table1")
-library("readxl")
-library("readr")
-library("dplyr")
-library("ggplot2")
-library("tidyverse")
-library("here")
+library(readxl)
+library(readr)
+library(dplyr)
+library(ggplot2)
+library(tidyverse)
+library(here)
+library(table1)
+library(plotly)
+library(ggtext)
+library(broom)
 # load the data sets
 Immunology_lab_results <- read_excel("data/raw-data/Immunology lab results.xlsx")
 View(Immunology_lab_results)
@@ -18,7 +22,7 @@ View(GENDER_DEUTERIUM_DILUTION_DATA)
 #select variables in both the data sets and later merge them together.
 
 gender_data<-GENDER_DEUTERIUM_DILUTION_DATA%>%
-  select(`Participant ID`, Sex, `Participant age`,BMI, Dose,`Fat in kg`,`LBM in kg`,
+  dplyr::select(`Participant ID`, Sex, `Participant age`,BMI, Dose,`Fat in kg`,`LBM in kg`,
           `Fat %`, `LBM in %`)
          
 gender_data<-gender_data%>%
@@ -26,7 +30,7 @@ gender_data<-gender_data%>%
 
 #select variables you will need in the next data set
 immune_data<-Immunology_lab_results%>%
-  select(ID,`CD4+`, `CD4 Immune activation count`)
+  dplyr::select(ID,`CD4+`, `CD4 Immune activation count`)
 
 #we can now merge the two data sets int one.
 Tb_immune_data<-full_join(gender_data, immune_data,by = "ID")
@@ -179,18 +183,20 @@ ggplot(Final_data, aes(x = factor(Sex), y = `CD4 Immune activation count`, color
 ######making a table with sex as our exposure.
 #installing the required packages for the tables.
 # Load required packages
-install.packages("table1")  # Install if not already installed
-install.packages("gtsummary")
-library(table1)
-library(gtsummary)
-library(plotly)
-library(ggtext)
+#install.packages("table1")  # Install if not already installed
+#install.packages("gtsummary")
+
 #descriptive statistics table with Sex as the exposure:
 TABLE1<-table1(~ `CD4+`+`CD4 Immune activation count` + `Participant age` + `Fat in kg` 
        + `LBM in kg` | Sex, data = Final_data)
 TABLE1
 save_tabel1<- here::here("results","tables", "TABLE1.rds")
 saveRDS(TABLE1,file =save_tabel1 )
+
+
+#table version 2
[…]

### 2025-04-05 products/manuscript/Manuscript.qmd
commit all changes  made on the code and manuscript and products

@@
 
-we shall use data from a tuberculosis public repository.
+#Methods
+
+We shall use data from a tuberculosis public repository.
 •	Data source: Tidyverse 
 https://dataverse.unc.edu/dataset.xhtml?persistentId=doi:10.15139/S3/AYOFEU
-with a Sample size of 60 individuals. The data includes 60 different individual that accepted to take part in the study.
+with a Sample size of 60 individuals. The study had 30 males and 30 females. All these participants were hiv negative and didnot have tuberculosis at the time of enrollment.
 
-The key variables that we shall look at are;
-
+The key exposure varibles are demographic and clinicalk. These include;
 Key variables: 
@@
 - Age of the participant
+
+#Proposed Analysis Method.
+
+•	The annalysis was conducted in R version 4.4.2. This was done to Compare immune activation levels between males and females.
+
+ We computed the male-to-female differences in disease severity using cd4 immune activation counts.We correlated immune markers with disease severity measures/Correlation between activation levels and disease severity.
+We did this using linear regression and machine learning models to predict which sex will show the most results.
+We also adjusted for potential confounders like age.
+
+#Results
+
+#conclusions.
+Sex differences in TB is contributed by more than just hormonal influence.
+Our study shows that TB is more severe in men than it is in women.
+Regardless of the medical seeking behaviors of women, women tend to a have a high body fat percentage more than men.The higher the fat percentage an individual has, the lower the risk of progression of the tuberculosis infection to disease hence lowering mortality rates among females compared to men.
+In addition, women tend to have more lean body mass compared to men hence having a lower chance of suffering from adverse effects of TB compared to men.
+
 ```{r}
@@
 ```
-
-Proposed Analysis Method.
-
-•	Compare immune activation levels between males and females
-
-•	Correlate immune markers with disease severity measures/Correlation between activation levels and disease severity
-
-•	Use linear regression to analyze relationships between variables
-
-•	Control for potential confounding factors
-
-These are some of our Expected Outcomes.
-
-•	Check and Identify any sex-based differences in immune response.
-
-•	Quantify the relationship between immune activation and disease severity.
-
-•	Propose any insights for sex-specific treatment approaches
 
@@
 
+We applied three supervised machine learning models — linear regression, LASSO regression, and Random Forests — to predict CD4 Immune Activation Count based on Sex, Age, BMI, Lean Body Mass (LBM), Fat mass, and CD4+ cell count.
+Data preprocessing included median imputation for missing values, dummy encoding of categorical variables, normalization, and removal of highly correlated features.
+Models were evaluated using RMSE and visualized through observed vs.
+predicted plots.
+Hyperparameter tuning was performed using grid search with cross-validation.
+
+The significance of Sex as a predictor of immune activation was assessed in each model.
+For linear models, coefficients for Sex were examined along with their corresponding p-values.
+The Random Forest model’s variable importance was assessed to identify the predictive contribution of Sex.
+
 {{< pagebreak >}}
@@
 To improve the model, we added in other predictors that could influence the immune activation count by performing a multi-variant linear regression anlaysis.
-This was done make sure we minimize confounding.
 
@@
 ```
+We finally considered using ML algorithm to fit our data.
+The Random Forest model achieved the lowest RMSE (1132), outperforming both the linear regression and LASSO models, which had RMSEs around 2107.
+This figure @Predicted_vs_observed shows a comparison of observed vs.
+predicted values for each model.
 
-We went throught the steps of model building.
-First, we tested the null model which comprised of our main outcome and sex as the predictor.
-We did a stepwise selection using a foward direction approach with a our full model with all the interaction effects.
-We found that adding the interaction terms did not improve the model performance.
-We then considered the null model in comparison with the full model but this time round without interaction effects but all the predictors.
-Using the same procedure of stepwise selection using the foward direction chose the a simple linear model with just cd4+ as the only additional predictor.
-We performed a simple linear regression on the data between the CD4 immune activation count as the main outcome and sex plus cd4+ as predictors. We considered females  to be our reference level.
-
-This @tbl-model_table3 shows a summary of a  final linear model fit.
 ```{r}
-#| label: tbl-model_table3
-#| tbl-cap: "Linear model fit table 3."
+#| label: model-1
+#| fig-cap: "predicted vs observed of three models"
 #| echo: FALSE
-model_table3 = readRDS(here("results","tables","model_table3.rds"))
-knitr::kable(model_table3)
+knitr::include_graphics(here("results","figures","predVSobs.png"))
 ```
+LASSO did not substantially improve performance over linear regression, likely due to a weak regularization effect at the selected penalty.
+Variable importance from the Random Forest model identified BMI, participants age,lean body mass,CD4+ and sex, with Sex emerging as a minor contributor to CD4 Immune Activation count levels.
 
@@
 
+•	Check and Identify any sex-based differences in immune response.
+
+•	Quantify the relationship between immune activation and disease severity.
+
+•	Propose any insights for sex-specific treatment approaches
+
+Despite limitations of the present study, this study suggests that the body regains weight among wasted tuberculosis patients.
+Patients with wasting do not recover to the same levels of body mass as those who were not wasted.
+This suggests that tuberculosis leads to permanent loss of lean tissue and fat mass.
+There was remarkable gender but not HIV differences in longitudinal body composition changes during the initial phase of tuberculosis treatment among patients who presented with body wasting.
+The substantial body composition differences between men and women suggest that the nutritional demands during recovery may differ for men and women.
+Evaluation of nutritional status should involve evaluation of lean tissue and fat mass and further studies should explore the impact of providing nutritional interventions as adjuvant treatment on body composition among tuberculosis patients during and after treatment in sub-Saharan Africa.
 {{< pagebreak >}}

### 2025-04-05 code/analysis-code/statistical-analysis.R

@@ -64,7 +64,6 @@ Final_data <- data %>%
   ))
 
 
-
 ##eXPLORATION OF OUR DATA.
 #summary of the data
 summary(Final_data)
@@ -203,16 +202,7 @@ saveRDS(TABLE1,file =save_tabel1 )
 t.test(`CD4 Immune activation count` ~ Sex, data = Final_data)
 t.test(`Fat in kg`~ Sex, data = Final_data)
 
-#Final_data %>%
- # select(Sex,`CD4+`, `CD4 Immune activation count`, `Participant age`, 
-   #      `Fat in kg`, `LBM in kg`) %>%
-  #tbl_summary(by = Sex, 
-   ##           statistic = list(all_continuous() ~ "{mean} ({sd})"), 
-     #         missing = "no") %>%
-  #add_p() %>%
-  #modify_header(label = "**Variable**") %>%
-  #modify_spanning_header(c("stat_1", "stat_2") ~ "**Sex Groups**") 
-#bold_labels()
+
 
 #simple models with one predictor.
 #since sex is our major exposure of interst, we shall consider it first and then 
@@ -266,11 +256,13 @@ library(MASS) # we shall need this package to perform a step aic for model selec
 forward_model <- stepAIC(null_model, 
                       scope = list(lower = null_model, upper = model4),
                       direction = "forward")
+#we maintained the sex variable in the null model because it our main predictor and we dont
+#want to lose it.
 
 # View the summary of the selected model
 summary(forward_model)
 
-#what if i wanted to tale into account the interaction effects
+#what if i wanted to take into account the interaction effects
 # Define full model with interactions
 full_model_with_interactions <- lm(`CD4 Immune activation count` ~ 
                                      (relevel(factor(Sex), ref = "F") + 
@@ -298,4 +290,153 @@ knitr::kable(model_table3)
 
 save_table3<-here::here("results","tables", "model_table3.rds")
 saveRDS(model_table3,file =save_table3 )
+
+#lets try using the ml model_table1. load the necessary packages.
+library(glmnet)       # For LASSO regression
+library(ranger)
+library(recipes)
+library(tidymodels)  
+library(car)
+# . Set random seed for reproducibility
+set.seed(1234)
+rngseed <- 1234
+
+# Fix column names
[…]

### 2025-04-05 products/manuscript/Manuscript.qmd
commit to the changes in the files

@@
 #| echo: FALSE
-knitr::include_graphics(here("results","figures","predVSobs.png"))
+knitr::include_graphics(here("results","figures","Predicted_vs_observed.png"))
 ```

### 2025-04-08 products/manuscript/Manuscript.qmd
save all the changes in the manuscript

--- 2025-04-05 products/manuscript/Manuscript.qmd
@@
 
+This figure @Variable_plot shows the variable importance plot from the random forest model.
+This highlights  the most influential predictors of CD4 immune activation.
+BMI, participants age, CD4+, Lean body mass(LBM) and Sex emerged as the most important features, indicating they contributed most to the model's predictive accuracy.
+This suggests that body composition and biological sex may play a key role in immune activation patterns.
+
 ```{r}
@@
 #| echo: FALSE
-knitr::include_graphics(here("results","figures","predVSobs.png"))
+knitr::include_graphics(here("results","figures","Predicted_vs_observed.png"))
 ```
-LASSO did not substantially improve performance over linear regression, likely due to a weak regularization effect at the selected penalty.
-Variable importance from the Random Forest model identified BMI, participants age,lean body mass,CD4+ and sex, with Sex emerging as a minor contributor to CD4 Immune Activation count levels.
+
+```{r}
+#| label: Forests-model
+#| fig-cap: "Variable plot from the random forests model."
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","Variable_plot.png"))
+```
+
+We applied LASSO regression to identify preredictors of CD4 immune activation.
+After removing the participant ID variable, the model retained BMI (estimate = -297), CD4+ count (estimate = 514), sex (male) (estimate = 260), and lean body mass (estimate = -22.7) as non-zero coefficients.
+Participant age was excluded from the final model.
+
+Interestingly, LBM (Lean Body Mass) and Participant Age were ranked lower, suggesting they had less influence in this dataset.
+These findings align with the Random forest results, which also retained BMI,CD4+ lean body mass and sex as key predictors.
 
@@
 # Discussion
+In this analysis of sex-based differences in TB-related immune activation, we aimed to identify key predictors of CD4 immune activation count , which we used as a proxy for severity of disease.
+Our hypothesis was that higher immune activation counts indicate more severe illness, and that biological sex may influence immune responses to TB.
 
+when we used both the LASSO regression and random forest models, we identified CD4+ count, BMI, sex, and lean body mass (LBM) as important predictors of immune activation.
 Our findings highlight a notable difference in CD4 immune activation count rates between males and females, with males exhibiting higher risk of TB infection and disease than females.
-The results show that men have a 735.34 increased immune activation count as compared to women.
+Sex was retained as a significant predictor, with a positive coefficient (estimate = 260).
+This indicated that males experienced a 260 increase in  CD4 immune activation count compared to females, who served as the reference group.
+This suggests that males may experience more severe immune activation, supporting the hypothesis that sex-based biological differences contribute to disease severity in TB.
 This result aligns with previous studies that have reported sex-based disparities in health outcomes and mortality risks as a result of TB[@Hertz2018],[@Mara2018].
@@
 
-## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
+## Strengths and Limitations
+This study has several strengths.
+We applied simple linear regression, LASSO regression and random forest models for annalysis of our data.
+We were able to capture a range of relationships between predictors and immune activation, which in turn increased the robustness of our findings.
+The focus on sex-based differences in TB immune response addressed an important and often neglected question in infectious disease research especially among TB patients who are HIV negative.
+Additionally, the use of biologically meaningful predictors such as BMI, lean body mass, CD4+ count, and sex enhances the interpretability and relevance of the results.
+The consistency of key predictors across the LASSO model and the random forests model further supports the reliability of our conclusions.
+However, there are limitations to consider.
+The relatively small sample size may reduce the generalizability and stability of the findings.
+As an observational study, we are limited to identifying associations rather than making causal claims.
+Important clinical variables, such as TB stage or treatment history, may not have been included, potentially introducing residual confounding.
+Moreover, while CD4 immune activation was used as our marker for disease severity, it may not fully capture the clinical picture.
 
-## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
+Despite limitations of the present of our study, some studies show that Patients with wasting do not recover to the same levels of body mass as those who were not wasted.
+This suggests that tuberculosis may leads to permanent loss of lean tissue and fat mass.
 
 ## Conclusions
-_What are the main take-home messages?_
 
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
+Our results support the hypothesis that higher immune activation is associated with greater disease severity, and demonstrate that males exhibit higher CD4 immune activation levels compared to females, indicating they may experience more intense immune responses and possibly more severe disease outcomes.
+These findings highlight the importance of considering sex-based biological differences in TB research and clinical management.
 
-•	Check and Identify any sex-based differences in immune response.
+In addition to sex, BMI and lean body mass emerged as important predictors, suggesting that body composition may influence the immune landscape in TB.
+These insights highlight the need for sex-specific approaches in TB monitoring and treatment, and point to CD4 immune activation as a potentially valuable biomarker for assessing disease severity.
+However, the need for future research is needed to validate these findings especially among larger populations  to explore the underlying sex mechanism and body composition differences in regard to immune activation.
 
-•	Quantify the relationship between immune activation and disease severity.
-
-•	Propose any insights for sex-specific treatment approaches
-
-Despite limitations of the present study, this study suggests that the body regains weight among wasted tuberculosis patients.
-Patients with wasting do not recover to the same levels of body mass as those who were not wasted.
-This suggests that tuberculosis leads to permanent loss of lean tissue and fat mass.
-There was remarkable gender but not HIV differences in longitudinal body composition changes during the initial phase of tuberculosis treatment among patients who presented with body wasting.
-The substantial body composition differences between men and women suggest that the nutritional demands during recovery may differ for men and women.
-Evaluation of nutritional status should involve evaluation of lean tissue and fat mass and further studies should explore the impact of providing nutritional interventions as adjuvant treatment on body composition among tuberculosis patients during and after treatment in sub-Saharan Africa.
 {{< pagebreak >}}

### 2025-04-08 code/analysis-code/statistical-analysis.R

@@ -308,6 +308,8 @@ colnames(Final_data) <- make.names(colnames(Final_data))
 Final_data$CD4.Immune.activation.count[is.na(Final_data$CD4.Immune.activation.count)] <- 
   median(Final_data$CD4.Immune.activation.count, na.rm = TRUE)
 
+Final_data <- Final_data %>% dplyr::select(-ID)
+
 #we have decided to impute them since this individual has other values that are
 #important in our annalysis.
 

### 2025-04-18 products/manuscript/Manuscript.qmd
commit to all the changes made on the
manuscript and other analyises

@@
 
-This project is aimed at looking at the differences in the Tuberculosis immune activation between males and females.( Sex-based analysis/Differences in TB Immune Activation
-Research Question)
+Tuberculosis is one of the leading causes of death in the world.According to NIH reports, tuberculosis (TB), afflicts more human males than females, with a ratio of 1.7 males to female.
+The disease is still endemic in most of the sub-Saharan African countries.
+In Uganda, for every one case of tuberculosis reported, there are at least three cases of tuberculosis that are not identified and its because of these cases that the disease remains prevalent in the country.
+Previous studies show that there are differences in the immune activation of TB among men and women.
+In this study, we aim to explore possible underlying hormonal and genetic mechanisms that might contribute to differences in the immune activation among females and males.
 
 #Methods
+We utilized data from a tuberculosis public repository available through Tidyverse (https://dataverse.unc.edu/dataset.xhtml?persistentId=doi:10.15139/S3/AYOFEU).
+The dataset comprises 60 individuals - 30 males and 30 females.
+All participants were HIV-negative and tuberculosis-free at the time of enrollment.
 
-We shall use data from a tuberculosis public repository.
-•	Data source: Tidyverse 
-https://dataverse.unc.edu/dataset.xhtml?persistentId=doi:10.15139/S3/AYOFEU
-with a Sample size of 60 individuals. The study had 30 males and 30 females. All these participants were hiv negative and didnot have tuberculosis at the time of enrollment.
+The key variables in our analysis include:
 
-The key exposure varibles are demographic and clinicalk. These include;
-Key variables: 
-- CD4 Immune activation count
--	Disease severity indicators(Extent of lung disease andNumber of lungs involved)
--Fat in kgs
-- Sex/Gender
-- BMI
-- Age of the participant
+CD4 immune activation count
+Disease severity indicators (extent of lung disease and number of lungs involved)
+Fat in kilograms
+Sex/Gender
+BMI
+Age of participant
+These demographic and clinical variables form the foundation of our investigation into tuberculosis-related immune responses.
 
@@
 
-•	The annalysis was conducted in R version 4.4.2. This was done to Compare immune activation levels between males and females.
+The annalysis was conducted in R version 4.4.2. This was done to Compare immune activation levels between males and females.
 
@@
 #Results
+# Results
+
+Our analysis revealed a statistically significant difference in CD4 immune activation counts between males and females.
+On average, females showed lower CD4 activation counts than males (mean difference = 413) indicating less immune activation in females and a true difference in the body fat between females and males (mean difference = 17.76 95%CI (7.097298 29.972199).
+
+Linear regression models adjusted for age, BMI, and fat mass demonstrated that male sex was independently associated with higher CD4 immune activation (β =  692.58 95% CI = [-786.97, 2172.13]).
+
+Disease severity indicators were more strongly associated with immune activation in males than in females, suggesting a sex-based difference in how immune response relates to disease burden.
+
+The LASSO and Random Forest models identified BMI, sex, CD4+, LBM  as the most important predictors of immune activation.
+These findings support the hypothesis that sex-based physiological differences, particularly in fat composition, contribute to variability in immune activation and disease severity in tuberculosis.
 
@@
 #summary of the data
-summary(data)
+#summary(data)
 ```
@@
 
-The World Health Organisations, among others, attributed to the differences in the tuberculosis trends between men and women to access to care and treatment, pregnancy effects,body fat mass and gender rules that diminish the social capital of women hence lowering incidence among them[@CCWhalen2023].
-Not forgetting that biological mechanisms might influence these differences,  previous studies show that men have a higher incidence and prevalence of disease and this might lead to bad treatment outcomes as compared to women.
+The World Health Organization, among other institutions, attributes differences in tuberculosis trends between men and women to several factors: access to care and treatment, pregnancy effects, body fat mass, and gender roles that diminish women's social capital, thereby lowering incidence among them [@CCWhalen2023].
+Worldwide, epidemiological data indicate that tuberculosis (TB) rates among HIV-negative men have consistently been higher than among women for many decades [@Horton2016].
+While biological mechanisms likely influence these differences, previous studies demonstrate that men experience higher incidence and prevalence of disease, potentially leading to poorer treatment outcomes compared to women.
+According to an analysis conducted in 2000, TB prevalence among males exceeded that among females in 27 (93%) of 29 prevalence surveys conducted across 14 countries between 1953 and 1997 [@Horton2016].
 
-Body fat plays an important role in supporting immune function, and women typically maintain higher fat storage than men.
+Other studies have highlighted that body fat plays an important role in supporting immune function, with women typically maintaining higher fat storage than men.
 This increased fat storage may contribute to women's stronger immune responses.
-Since the balance of fat and lean tissue differs significantly between men and women, particularly at the time of disease presentation, the objective of this study is to determine whether higher immune activation correlates with greater infection severity or illness.
+While these differences are partly attributed to women's better healthcare-seeking behaviors, factors such as fat distribution, hormonal profiles, body composition, lean tissue mass, and other biological characteristics differ significantly between sexes, particularly at disease presentation.
+The objective of this study is to determine whether higher immune activation counts correlate with greater infection severity or illness progression.
 
@@
 ## Schematic of workflow
-Chromosomal sex and hormones play a role in the modulation of immune responses to Mtb and thereby contribute to the general outcome of disease.
+Chromosomal sex and hormones play a role in the modulation of immune responses to mTB and thereby contribute to the general outcome of disease.
 The first cells that encounter Mtb are known as Macrophages.
@@
 
-We explored the data and using figures and summary tables.
-We used box plots, histograms and and correlation plots to see whether we can find something interesting in the data.
-Scatter plots were used to see the trends in the data and also see whether there are some interesting relationships between the continuous variables in the data.
+We conducted a thorough data exploration using various visualization techniques and summary tables.
+Our analysis included box plots, histograms, and correlation plots to identify meaningful patterns in the dataset.
+We employed scatter plots to examine trends and investigate potential relationships between continuous variables.
+This comprehensive approach allowed us to uncover several interesting insights within the data.
 
-@tbl-TABLE1 shows a summary of the data
+@Table1 shows the summary statistics of our data variables.
+```{r}
+# Load the table
+#table1 <- readRDS(here::here("results", "tables", "TABLE1.rds"))
+#| label: table-Table1
+#| tbl-cap: "Summary distribution of baseline characteristics among males and females."
+#| echo: FALSE
 
-```{r}
-#| label: tbl-TABLE1
-#| tbl-cap: "Data descriptive summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/tables/TABLE1.rds")
-knitr::kable(resulttable)
+# Display the table (use knitr::kable, gt, or another display function)
+knitr::kable(table1)
+
 ```
@@
 ## Basic statistical analysis
-We used simple linear model to inspect whether there was an association between our main outcome CD4 immune activation count and the the primary predictor sex.
-To improve the model, we added in other predictors that could influence the immune activation count by performing a multi-variant linear regression anlaysis.
 
-@fig1-result shows a scatter plot of lean body mass in kg and CD4
+We employed a simple linear model to investigate the potential association between our primary outcome, CD4 immune activation count, and our main predictor variable, sex.
+To enhance the model's explanatory power, we incorporated additional predictors that might influence immune activation counts through multivariate linear regression analysis.
 
-```{r}
-#| label: fig1-result
-#| fig-cap: "Scatter plot of CD4+ cells versus LBM in kg."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","scatterplot1.png"))
-```
-
-@fig2-result shows a scatter plot of CD4 vs Fat in kg.
-```{r}
-#| label: fig2-result
-#| fig-cap: "Scatter plot of CD4+ cells versus Fat in kg."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","scatterplot2.png"))
-```
-
-@plot1-result a boxplot of the cd4 immune activation count by sex.
-```{r}
-#| label: plot1-result
-#| fig-cap: "Box plot to show the distribution of cd4 count immune activation counts among males and females."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","boxplot1.png"))
-```
-
-@plot2-result a boxplot to show the distributions of fat in kilograms among males females.
+Given that fat content serves as a contributing factor to immune activation in disease progression, individuals with higher fat percentages typically demonstrate reduced susceptibility to tuberculosis, and vice versa.
+We analyzed the distribution of fat composition between men and women.
+Figure @plot2-result presents a boxplot illustrating the distribution of fat in kilograms across male and female participants.
 ```{r}
@@
 
-This @tbl-model_table2 shows a summary of a  multi-variant linear regression model fit.
-```{r}
-#| label: tbl-model_table2
-#| tbl-cap: "Linear model summary table 2."
-#| echo: FALSE
-model_table2= readRDS(here("results","tables","model_table2.rds"))
-knitr::kable(model_table2)
-```
 We finally considered using ML algorithm to fit our data.
@@
 
-We applied LASSO regression to identify preredictors of CD4 immune activation.
+We applied LASSO regression to identify predictors of CD4 immune activation.
 After removing the participant ID variable, the model retained BMI (estimate = -297), CD4+ count (estimate = 514), sex (male) (estimate = 260), and lean body mass (estimate = -22.7) as non-zero coefficients.

### 2025-04-18 code/analysis-code/statistical-analysis.R

@@ -214,10 +214,12 @@ model2 <- lm(`CD4+` ~ relevel(factor(Sex), ref = "M"), data = Final_data)
 summary(model2)
 
 #Using cd4 immune activation count .
-model3 <- lm(`CD4 Immune activation count` ~ relevel(factor(Sex), ref = "F"), data = Final_data)
+model3 <- lm(`CD4 Immune activation count` ~ relevel(factor(Sex), ref = "F") + 
+               BMI + `LBM in kg`  , data = Final_data)
 
 model_summary <- summary(model3)
 model_summary
+confint(model3)
 model_table1 <- as.data.frame(model_summary$coefficients)
 knitr::kable(model_table1)
 

### 2025-04-24 products/manuscript/Manuscript.qmd
commit to the changes

@@
 
-```{r}
-#load packages
-library("readxl")
-library("readr")
-library("dplyr")
-library("ggplot2")
-library("tidyverse")
-library("here")
-
-# follow through with the code folder to see the #variables in the data and how i selected the #variables make this data set. 
-#location of the data
-data_location <- here::here("data","processed-data","Tb_immune_data.rds")
-
-#loading the data
-data<- read_rds(data_location)
-
-#summary of the data
-#summary(data)
-```
-
 {{< pagebreak >}}
@@
 This is considered a safe place where the the Tuberculosis data from the TRAc project is kept and can be used by anyone who wants contribute to TB research.
-More details about the data can be found [here]!(https://dataverse.unc.edu/dataset.xhtml?persistentId=doi:10.15139/S3/AYOFEU).
-The data is publicaly available and can be accessed by anyone.
+More details about the data can be found [here](https://dataverse.unc.edu/dataset.xhtml?persistentId=doi:10.15139/S3/AYOFEU).
+The data is publicly available and can be accessed by anyone.
 
@@
 
-\textbf{inclusion and exclusion critera}
+###inclusion and exclusion critera
 
-Our study included patients aged 15 years and older who were experiencing their first episode of tuberculosis.
-To maintain data integrity, we excluded individuals who were receiving hormonal therapy or using hormonal contraceptives.
-Additionally, patients with comorbidities such as diabetes mellitus, cancer, or HIV were excluded from the study.
-Conversely, we applied strict exclusion criteria to ensure a focused research population: all individuals under 15 years of age, those with recurrent TB episodes, patients taking any form of hormonal therapy or contraception, and those diagnosed with diabetes mellitus, cancer, or HIV were not enrolled in the study.
+Our study included patients aged 15 years and older experiencing their first episode of tuberculosis.
+To maintain data integrity, we excluded individuals receiving hormonal therapy or using hormonal contraceptives.
+Additionally, patients with comorbidities including diabetes mellitus, cancer, or HIV were excluded.
+Our strict exclusion criteria ensured a focused research population by omitting all individuals under 15 years of age, those with recurrent TB episodes, patients taking any form of hormonal therapy or contraception, and those diagnosed with diabetes mellitus, cancer, or HIV.
 
 The variables of interest were extracted from the TRAC dataset  and exported to an MS Excel sheet.
-The outcome variable was CD4+ immune activatin counts.
+The outcome variable was CD4+ immune activation counts.
 The primary independent variable was sex (cis-gender) and this was categorized as male and female.
@@
 This causes the long term containment of TB infection to differ between men and women since their adaptive immune responses are different.
- @tbfig-1
+See @tbfig-1
 
@@
 ```{r}
-# Load the table
-#table1 <- readRDS(here::here("results", "tables", "TABLE1.rds"))
-#| label: table-Table1
+#| label: tbl-table1
 #| tbl-cap: "Summary distribution of baseline characteristics among males and females."
 #| echo: FALSE
-
-# Display the table (use knitr::kable, gt, or another display function)
-knitr::kable(table1)
+table1<-readRDS(here::here("results", "tables", "TABLE1.rds"))
+table1
 
 ```
+
+Table @tbl-correlation shows the correlation matrix between immune activation markers and body composition variables.
+
+```{r}
+#| label: tbl-correlation
+#| tbl-cap: "Correlation matrix between immune activation markers and body composition variables."
+#| echo: FALSE
+
+# Load the table
+cor_table <- readRDS(here::here("results", "tables", "TABLE2_Correlation.rds"))
+
+cor_table
+
+```
+Table 2 presents the Pearson correlation coefficients between CD4 immune markers and body composition variables.
+CD4+ count showed a modest positive correlation with both CD4 immune activation count (r = 0.22) and fat mass (r = 0.17), and a negative correlation with lean body mass (r = –0.18).
+Notably, fat in kilograms was strongly negatively correlated with lean body mass (r = –0.94), indicating an inverse relationship between fat and lean tissue.
+Participant age showed very weak correlations with all variables, suggesting age may not substantially confound relationships between immune activation and body composition in this sample.
 
@@
 We analyzed the distribution of fat composition between men and women.
-Figure @plot2-result presents a boxplot illustrating the distribution of fat in kilograms across male and female participants.
+This @plot2-result presents a boxplot illustrating the distribution of fat in kilograms across male and female participants.
 ```{r}
@@
 ## Full analysis
-In order to take care of confounding, we performed a multi-regression analysis and the results of the analysis are summarized in the table below.
+In order to take care of confounding, we performed a multi-regression analysis and the results of the analysis were not satifactory.
 

### 2025-04-24 code/analysis-code/statistical-analysis.R

@@ -10,6 +10,7 @@ library(table1)
 library(plotly)
 library(ggtext)
 library(broom)
+library(flextable)
 # load the data sets
 Immunology_lab_results <- read_excel("data/raw-data/Immunology lab results.xlsx")
 View(Immunology_lab_results)
@@ -102,6 +103,11 @@ print(cor_matrix)
 corrplot(cor_matrix, method="color", tl.cex=0.8)
 #We see that none of the variabls are correlated with each other.
 
+library(ggplot2)
+library(GGally)
+ggpairs(Final_data[, c("CD4+", "Fat in kg", "BMI", "Participant age")])
+
+
 #Scatter plots of numeric variables
 p1<-ggplot(Final_data, aes(x=`LBM in kg`
                  , y=`CD4+`)) + 
@@ -193,8 +199,31 @@ save_tabel1<- here::here("results","tables", "TABLE1.rds")
 saveRDS(TABLE1,file =save_tabel1 )
 
 
-#table version 2
-# ========= TABLE 1: Demographics and Clinical Characteristics by Sex =========
+
+#table version 2, correlation table
+# Load necessary packages
+
+library(gt)
+library(tibble)
+# Select numeric variables
+cor_data <- Final_data %>%
+  select(`CD4+`, `CD4 Immune activation count`, `Fat in kg`, `LBM in kg`, `Participant age`, BMI)
+
+# Compute correlation matrix
+cor_matrix <- round(cor(cor_data, use = "complete.obs"), 2)
+
+# Convert to a data frame and make it prettier
+cor_df <- as.data.frame(cor_matrix) %>%
+  rownames_to_column("Variable")
+
+# Format as a gt table
+cor_table <- gt(cor_df) %>%
+  tab_header(title = "Table 2: Correlation Matrix of Immune Activation and Body Composition")
+cor_table
+# Save as RDS for use in Quarto
+saveRDS(cor_table, file = here("results", "tables", "TABLE2_Correlation.rds"))
+
+
 
 
 

### 2025-04-24 project-review-Yufei Wu.qmd

---
title: Project Review_Yufei Wu
author: Yufei Wu
date: 04/23/2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Sex-based analysis/Differences in TB Immune Activation

Name of project author(s): Doreen Kibuule Kalembe

Name of project reviewer: Yufei Wu

# Specific project content evaluation

## Background, Context and Motivation

### Feedback and Comments

The background part provides a thorough overview of the TB burden and sex differences, including a summary of previous work. At the end of the introduction, it may be good to make a transition to the objective of this study by stating the what is still unknown and why this study addresses it.

### Summary assessment

* strong contextualization and motivation

## Question description

### Feedback and Comments

The question the project aims to address is clearly described.

### Summary assessment

* question/hypotheses fully clear

## Data description

### Feedback and Comments

The data source is well provided. It would be better if the codebook were also provided.

### Summary assessment

* source and overall structure of data somewhat explained

## Data wrangling and exploratory analysis

### Feedback and Comments

The data cleaning part looks good. For the exploratory analysis, it says that box plots, histograms, and correlation plots were included, but these are not shown in the manuscript or the supplementary information, just in the figure file. It’d be clearer if those plots were actually included in the manuscript or supplement. In addition, it would strengthen the analysis if more extensive EDA were conducted (if possible).

### Summary assessment

* some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

### Feedback and Comments

Analysis methods are appropriate. It will be better to provide more cross-validation details, like the CV RMSE.

### Summary assessment

* strong and reasonable analysis

## Presentation

### Feedback and Comments

The overall presentation is good. However, there' some room for improvement. Some figures are mentioned but not included (like the exploratory plots). It's better to remove the R warning consoles. In addition, the links to figures and tables are not working properly.

### Summary assessment

* results are presented ok, with room for improvement

## Discussion/Conclusions

### Feedback and Comments

The discussion is detailed and includes proper interpretations of findings. Strengths and limitations are also discussed.

### Summary assessment

* strong, complete and clear discussion

## Further comments

The manuscript will be better if it is polished to be neater. Also, the spelling may be checked (like “annalysis” → analysis, “activatin” → activation). The title of the manuscript "Example Manuscript Template for a Data Analysis Project" should be replaced to the actual title.

# Overall project content evaluation

## Structure

### Feedback and Comments

The project is mostly well structured. However, there are some "junk" template pictures in the assets.

### Summary assessment

* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation

### Feedback and Comments

The project is well documented and I can understand the analysis.

### Summary assessment

* fully and well documented

## Reproducibility

### Feedback and Comments

The results are mostly reproducible, but I did run into a minor issue. In the "statistical-analysis.R" file, line 48 reads data <- read_rds(data_location), but it seems like it should be save_data_location instead of data_location. I got an error when I first ran the analysis file on its own. However, if I ran the manuscript file first and then went back to the analysis script, it worked—likely because the data was already loaded from the manuscript file.

### Summary assessment

* small parts not reproducible or required manual intervention

## Thoroughness

### Feedback and Comments

Different models are considered and the hypothesis is addressed.

### Summary assessment

* strong level of thorougness

## Further comments

The overall project is complete and understandable. The manuscript file may need to be further polished to meet publication standard.

### 2025-04-24 products/manuscript/Manuscript.qmd
stage all changes to the files

@@
 ---
-title: "Example Manuscript Template for a Data Analysis Project"
+title: "Sex-based analysis/Differences in TB Immune Activation"
 author: "Doreen Kibuule Kalembe"

### 2025-04-24 code/analysis-code/statistical-analysis.R

@@ -46,7 +46,7 @@ saveRDS(Tb_immune_data, file = save_data_location)
 #dropping the last observation from the dataset since the last observation
 #doesnt make sense to our annalysis 
 #loading the data
-data<- read_rds(data_location)
+data<- read_rds(save_data_location)
 
 # dropping the last observation in the dataset since it makes no sense
 # to our annalysis and also dropping the variable dose since we dont really need it.

### 2025-04-27 project-review-prasanga/project-review-prasanga.qmd

---
title: Project Review Template 
author: Prasanga Paudel
date: April 23, 2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Sex-based analysis/Differences in TB Immune Activation

Name of project author(s): Doreen Kibuule Kalembe

Name of project reviewer: Prasanga Paudel

# Specific project content evaluation
We will evaluate the different parts of the project  in the sections below.

## Background, Context and Motivation

How well is the context of the project described?

- There is a solid background section and it provides enough context of the project.

Is a comprehensive background, including summary of previous/related work given?

- Yes, they are very much related to the project.

Is the project well placed into the context of existing work (including proper referencing of existing work)?

- Yes, the references are impressive.

Is it clear why the project was undertaken and what new information it hopes to provide?

-  Yes, the information is provided but it could be more better if the the objectives and hypotheis are clearly listed.

### Feedback and Comments

The summary section of the manuscript has a lot of issues primarily related to the structure and contents. The overall content is very good, however, some warnings are popping up in the manuscript which should be fixed. The summary section also shows codes and outputs from R. The references are also not properly used. The hashtags in the Abstract section looks unprofessional.

Overall the information is of very good quality but it needs final refinement.

### Summary assessment

* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described?

- somewhat inconsistent objectives and hypothesis

Is it clear how the questions relate to the data?

- yes, it is somewhat clear.

### Feedback and Comments

I think the objective of the research and hypothesis are quite differently mentioned. The objective is stated as "the objective of this study will be to find out whether there are differences in the Tuberculosis immune activation between males and females" but the hypothesis is set as "We shall hypothesize on the more immune activation there is, the more infected / sicker the person". So there's some confusion in what the primary goal of the project is.

The author also mentioned in the introductory section that "The objective of this study is to determine whether higher immune activation counts correlate with greater infection severity or illness progression". I think there should be a clear objective and if there are multiple objectives, they should be listed in the objective section.

The data is not described in detail. There should be a brief description about what kinds of data can be expected in the dataset and about the time and location of the dataset of interest

### Summary assessment

* question/hypotheses somewhat explained

## Data description

How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

- The data source is provided but the variable description is missing. There is no codebook.

### Feedback and Comments

There are variables such as FMI, FFMI, Fat%, LBM, and TBW which require more explanation of what they stand for. A codebook would make things right. Variables should be describe well at the top or in a readme file.  Two datasets are merged at the beginning, and it would be more better if the reason for merging those datasets were clearly explained.

### Summary assessment

* source and overall structure of data somewhat explained

## Data wrangling and exploratory analysis

How well is the data cleaned/processed and explored?

- The dataset looks quite clean in the original dataset, hence little requirement for cleaning the data makes sense. The author merged two dataset which is quite impressive and reflects their motivation. Data exploration is somewhat weak. There are a lot of variables in the dataset which went un-noticed. Few scatterplots have been created which look interesting but surprisingly they have not been included in the manuscript.

Are all steps reasonable and well explained?

- The steps looks reasonable but there should have been more tables and figures that explore the basic relationship among factors of interest. And, more descriptions.

Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

- I could not find the supplementary materials. The codes do not create much figures other than those used in the manuscript and some scatterplots. I checked the figures folder but there are some unnecessary file related to weight and height distributions which I think should be removed. I could only see a table and a box-plot of male an female as EDA in the manuscript. The table has information on countries which I find to be very strange and unrelated.

### Feedback and Comments

The EDA section should include figures that could explain the relationship among various variables in the dataset. Figures such as scatterplot and correlation plot would provide a great insight into the data. The EDA should establish a story that could help us understand the analysis part of the project better. This is somewhat missing.

### Summary assessment

* major weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

Were the analysis methods appropriate for the data? Was the analysis done properly?

- The author does talk about various models: Linear regression, LASSO, and random Forest. The importance table from RF model is presented in the manuscript. But, I could not find the results from Linear regression and LASSO in the manuscript. There is a observed-predicted plot for LASSO and RF model but not for Linear Regression. The analysis seems to be done in detail but it's kind of messy and due to the lack of sections and sub-sections.

Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

- The above mentioned components are partly covered.

### Feedback and Comments

The author does talk about RMSE as a performance measure but it would be more better to compare the performance through a clear table or separately is a different subsection. The variables selection process should also be mentioned more clearly. There are a bunch of regression models and an attempt to control for confounder but the logic behind the model should also be mentioned and enough explanation should be provided for following a certain process.

### Summary assessment

* defensible but not optimal analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

- There is only one table and one figure in the manuscript and unfortunately both the table and figure references are not working. The table and figure are intuitive but the information obtained from figures and tables could be more helpful if more figures for more variables were used. I saw a descriptive analysis based on male and female in the analysis-code but suprisingly it was not included in the manuscript. Improvements can be done in this part of the project.

### Feedback and Comments

The presentation of the manuscript can be made better if few figures and tables are added in the manuscript. Currently, the "EDA" has a table and there's one figure under "Basic Statistical Analysis". EDA could talk more about how the responses under different interesting variables are distributed and BSA could talk about how different variables affect the target variable.

### Summary assessment

* results are presented ok, with room for improvement

## Discussion/Conclusions

Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

- Yes the findings are properly discussed. The strengths and limitations are also clearly discussed.

### Feedback and Comments

The Discussion section is well prepared. Findings from all three models can also be discussed in terms of performance and robustness in results.

### Summary assessment

* strong, complete and clear discussion

## Further comments

_In summary, both the code files and manuscript needs to be differentiated into distinct section to carry out different type of analysis. It would make the manuscript better. _

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure

Is the project well structured?

- not quite perfectly

Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed?

- Yes, files are in reasonable folders and with reasonable names.
- There are still junk files totally unrelated to the project.

 By just looking at files and folders, can you get an idea of how things fit together?

- Somewhat yes.

### Feedback and Comments

- The title is not well placed.
- The main headings of the manuscript are well structred. But, there should be different subsections for diffent models. For instance, different sections for different models.
- There should be a section that compares the performance.

### Summary assessment

* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation

How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

- The analysis is somewhat not well structured. There should have been a clear sections for each model to discuss the results and performance in the codes as well. As the information is provided as comments it is very difficult to navigate through the codes. Chunks would have been more helpful in presenting the codes. Even without the chunks, there should be some spacing between codes that are different from eachothers.

- All the code has been placed under the same 'statistical-analysis' file which is not a good idea. There should have been separate files for each purpose. The 'statistical-analysis.R' file creates the processsed-data and then saves it and then imports it again. This process just adds more codes into the file withput much help.

### Feedback and Comments

There should be a different code file for processing-code at least. EDA and analysis could be placed together if space and presented properly.

### Summary assessment

* decently documented with some gaps

## Reproducibility

Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

- The codes are reproducible. But, I had to edit the analysis-codes as the original code did not run due to incorrect reference of a dataframe. The manuscript code ran without interventions.

### Feedback and Comments

The reproducibility of the project is quite good. Adding instruction in the readme file about how to go through the project might help. But, the author has done quite a good job at making the project reproducible.

### Summary assessment

* small parts not reproducible or required manual intervention

## Thoroughness

How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

- The overall study was thorough but it could have been better. Alternatives were explored for the main models but the alternatives should have been formally compared through tables. The manuscript does not talk about cross-validation and train/test techniques were not used in the project which could have been very useful.

### Feedback and Comments

The authur should include a table that compares the model performances in the train dataset in general and the cross-validation datasets. It is already late to use the train/test technique as all the data has been already used, but if possible the author may get more data from the source itself. The alternative models should be thoroughly discussed based on the performance across different robustness procedures.

### Summary assessment

* decent level of thoroughness

## Further comments

_Most of the information obtained from the analysis section are important to the project and should be presented in the manuscript. This will make the project more detailed and increases it's credibility._

### 2025-04-28 products/manuscript/Manuscript.qmd
commit to all changes in the manuscript

@@
 
-$\land$ Corresponding author: kelvinbwambale8@gmail.com
+$\land$ Corresponding author: Doreen K. Kalembe
 
@@
 While these differences are partly attributed to women's better healthcare-seeking behaviors, factors such as fat distribution, hormonal profiles, body composition, lean tissue mass, and other biological characteristics differ significantly between sexes, particularly at disease presentation.
-The objective of this study is to determine whether higher immune activation counts correlate with greater infection severity or illness progression.
+The objective of this study is to determine whether higher immune activation counts correlate with greater infection severity or illness progression and whether this relationship differs between males and females.
 
@@
 Since the balance of fat and lean tissue differ in men and women especially at the time of tuberculosis presentation, the objective of this study will be to find out whether there are differences in the Tuberculosis immune activation between males and females.
-We shall hypothesize on the more immune activation there is, the more infected / sicker the person.
+We shall hypothesize on the more immune activation there is, the more infected / sicker the person and whether this relationship differs between males and females.
 
@@
 We applied three supervised machine learning models — linear regression, LASSO regression, and Random Forests — to predict CD4 Immune Activation Count based on Sex, Age, BMI, Lean Body Mass (LBM), Fat mass, and CD4+ cell count.
-Data preprocessing included median imputation for missing values, dummy encoding of categorical variables, normalization, and removal of highly correlated features.
+Data pre processing included median imputation for missing values, dummy encoding of categorical variables, normalization, and removal of highly correlated features.
 Models were evaluated using RMSE and visualized through observed vs.
@@
 
-Given that fat content serves as a contributing factor to immune activation in disease progression, individuals with higher fat percentages typically demonstrate reduced susceptibility to tuberculosis, and vice versa.
+Since fat content serves as a contributing factor to immune activation in disease progression, individuals with higher fat percentages typically demonstrate reduced susceptibility to tuberculosis, and vice versa.
 We analyzed the distribution of fat composition between men and women.
-This @plot2-result presents a boxplot illustrating the distribution of fat in kilograms across male and female participants.
+This @Fig2 presents a boxplot illustrating the distribution of fat in kilograms across male and female participants.
 ```{r}
-#| label: plot2-result
+#| label: Fig2
 #| fig-cap: "Box plot showing Fat in kg among males and females."
@@
 
+While this @Fig3 shows the distribution of CD4 among males and females.
+```{r}
+#| label: Fig3
+#| fig-cap: "Regression plot showing CD4+ among males and females."
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "Sex_Regplot.png"))
+
+```
+
 ## Full analysis
-In order to take care of confounding, we performed a multi-regression analysis and the results of the analysis were not satifactory.
+In order to take care of confounding, we performed a multi-regression analysis and the results of the analysis were not satisfactory.
 
 We finally considered using ML algorithm to fit our data.
-The Random Forest model achieved the lowest RMSE (1132), outperforming both the linear regression and LASSO models, which had RMSEs around 2107.
+The Random Forest model achieved the lowest RMSE (1984.8), outperforming both the linear regression and LASSO models, which had RMSEs 2123.8 and 2070.1 respectively.
 This figure @Predicted_vs_observed shows a comparison of observed vs.
@@
 
-when we used both the LASSO regression and random forest models, we identified CD4+ count, BMI, sex, and lean body mass (LBM) as important predictors of immune activation.
-Our findings highlight a notable difference in CD4 immune activation count rates between males and females, with males exhibiting higher risk of TB infection and disease than females.
+Both the LASSO regression and random forest models identified CD4+ count, BMI, sex, and lean body mass (LBM) as important predictors of immune activation.
+Our findings from the lasso model highlight that participant age was not selected, indicating it was not an important predictor after accounting for other variables.The results showed that Higher CD4 counts were associated with higher immune activation (estimate = 514), supporting our hypothesis that immune system status is closely tied to disease severity.
+The coefficient for lean body mass (LBM) was small (-22.7), indicating only a minor relationship.A notable difference in CD4 immune activation count rates between males and females, with males exhibiting higher risk of TB infection and disease than females.
 Sex was retained as a significant predictor, with a positive coefficient (estimate = 260).
 This indicated that males experienced a 260 increase in  CD4 immune activation count compared to females, who served as the reference group.
+
 This suggests that males may experience more severe immune activation, supporting the hypothesis that sex-based biological differences contribute to disease severity in TB.
@@
 The relatively small sample size may reduce the generalizability and stability of the findings.
-As an observational study, we are limited to identifying associations rather than making causal claims.
+
+As any observational study, we are limited to identifying associations rather than making causal claims.
 Important clinical variables, such as TB stage or treatment history, may not have been included, potentially introducing residual confounding.

### 2025-04-28 products/supplementary folder/supplement.qmd
commit to all changes in the manuscript

This is just a detailed document of what I did at every stage of the analysis.

Data cleaning.

We merged two data sets. The study was conducted in uganda and the variables say blood sample were got and different variables were got for the same people. for the variables collected, two data sets were got out of them. Our study utilised variables from both sets of data sets.

After the merging process, we then selected the variables that would be ideal for the king of project we were working on and dropped the unnecessary ones at this point.

Our final data set contained the following variables.

1. sex: This was the sex of the individual. This was male and female.

2. Participant Age: The age of the participant in years.

3. BMI: The body mass index of a person.

4. Fat in Kg. Measure of a participants body fat in kg and their corresponding percentages.

5. LBM in kg:Measure of a participants Lean Body Mass (LBM)in kg and its corresponding percentage.

6. CD4+: The number of CD4+ count cells an individual has.

Other abbreviations that may be important to note are,

 FFM: This represent the fat mass index of an individual.

 FFMI: This represents an individual's fat-free mass index

### 2025-04-28 code/analysis-code/statistical-analysis.R

@@ -78,34 +78,75 @@ table(Final_data$Sex) #There are 30 males and 30 females
 table(Final_data$Age_cat)
 
 ##VISUALIZING THE VARIABLES IN THE Final_data
-# creating a hists 
-hist(Final_data$`CD4 Immune activation count`, main="Histogram of CD4 Counts",
-     xlab="CD4 counts", col="blue") #histogram showing the cd4 counts
+# creating a histograms
 
-#CD4+
-ggplot(Final_data, aes(x=`CD4+`)) + geom_histogram(binwidth=10, fill="blue", 
-                                             color="black")
+hist1 <- ggplot(Final_data, aes(x = `CD4 Immune activation count`)) +
+  geom_histogram(fill = "blue", color = "black") +
+  labs(
+    title = "Histogram of CD4 Counts",
+    x = "CD4 counts",
+    y = "Frequency"
+  )
 
-#LBM IN KG
-ggplot(Final_data, aes(x=`LBM in kg`)) + geom_histogram(binwidth=10, fill="red", 
+hist1
+
+# Save the plot
+ggsave(filename = here("results", "figures", "histogram1.png"), plot = hist1)
+
+
+#Histogram of CD4+
+HISTCD4<-ggplot(Final_data, aes(x=`CD4+`)) + geom_histogram(binwidth=10, fill="blue", 
+                                          color="black")
+
+HISTCD4  
+
+
+# Save the plot
+ggsave(filename = here("results", "figures", "histogramCD4.png"), plot = HISTCD4)
+
+
+#HISTOGRAM OF LBM IN KG
+Hist3<-ggplot(Final_data, aes(x=`LBM in kg`)) + geom_histogram(binwidth=10, fill="red", 
                                              color="black")
-##sexDistributions
-ggplot(Final_data, aes(x=factor(Sex), y= `LBM in kg`)) +
+# Save the plot
+ggsave(filename = here("results", "figures", "histogram3.png"), plot = Hist3)
+
+
+
+##sex Distributions using boxplots
+
+sex_Bplot<- ggplot(Final_data, aes(x=factor(Sex), y= `LBM in kg`)) +
   geom_boxplot(fill="lightblue") +
   theme_minimal()
+sex_Bplot
+
+# Save the plot
+ggsave(filename = here("results", "figures", "Sex_boxplot1.png"), plot = sex_Bplot)
 
 ##checking for any correlations
-
+##we needed to check whether there are varibles that are strongly related to each
[…]
